Parallelization of Fast $\ell_1$-Minimization for Face Recognition

### 2011-06-05 main.tex
Applied fixes from Dan.  Sent to Allen.

\begin{abstract} Recently a family of promising face recognition algorithms
based on sparse representation and $\ell_1$-minimization ($\ell_1$-min) have been
developed.  These algorithms have not yet seen commercial application, 
largely due to higher computational cost compared to other traditional
algorithms. This paper studies techniques for leveraging the 
massive parallelism available in GPU and CPU hardware to accelerate
$\ell_1$-min based on augmented Lagrangian method (ALM) solvers. 
Due to the low arithmetic intensity of ALM solvers, both
architectures are bandwidth limited.  
with large problems running faster on the GPU, and small problems running faster
on the CPU.  
On both architectures, the proposed implementations dramatically out-perform
nieve library-based implementations. 
At the problem sizes most important for $\ell_1$-min based
face recognition, both architectures achieve comparable performance.
The source code of the
algorithms will be made available for peer evaluation.  \end{abstract}

@@

Among its many applications, $\ell_1$-min has been recently used to reformulate
image-based face recognition as a sparse representation problem
\cite{WrightJ2009-PAMI}.  If we stack the training images of $K$ subject
classes $(A_1\in\Re^{d\times n_1}, \cdots, A_K\in\Re^{d\times n_K})$ into the
columns of a matrix $A = [A_1, \cdots, A_K]\in\Re^{d\times n}$, given a new
query image in the same vector form $\bb\in\Re^d$, \emph{sparsity-based
classification} (SBC) solves the following minimization problem:
\begin{equation}
\min_{\xx, \ee} \| \xx \|_1 + \|\ee\|_1 \quad \subj \quad \bb = A \xx + \ee.
\label{eq:l1min_denoise}
\end{equation}
If the sparsest solutions for $\x$ and $\e$ are recovered, $\ee$ provides a
means to compensate for pixels that are corrupted due to occlusion of some part of the query
image, and the dominant nonzero coefficients of $\xx$ reveal the membership of
$\bb$ based on the training image labels associated with $A$.  Since $A$ ideally
contains many images per user taken under different illuminations, 
$A\x$ also acts as a linear illumination model for the test image $\bb$.

In this paper, we study parallelization of $\ell$-min on many-core CPU and GPU
architectures. Although $\ell_1$-min \eqref{eq:l1min} is a convex
program, conventional algorithms such as interior-point methods
\cite{ChenS2001-SIAM,TibshiraniR1996} are known to suffer from poor scalability
for large-scale applications such as face recognition. Recently, a number of
accelerated algorithms have been proposed that explicitly take advantage of
the special structure of $\ell_1$-min problems
\cite{LorisI2009,YangA2010-ICIP}. We investigate parallelization of a
state-of-the-art $\ell_1$-min solution based on the classical framework of
\emph{augmented Lagrangian methods} (ALM) \cite{BertsekasD2003,YangA2010-ICIP}.

While solving \eqref{eq:l1min_denoise} can achieve exceptional
recognition accuracy on public databases, its success also relies on the assumption
that the query image is well aligned with the training images. In
\cite{WagnerA2009-CVPR}, the framework of SBC was extended to iteratively align
a query image to each subject class individually. The face alignment algorithm solves
the following minimization problem:
\begin{equation}
\hat{\tau}_i = \arg\min_{\xx, \ee, \tau_i} \|\ee\|_1\quad \mbox{subj. to}\quad \tilde\bb\circ\tau_i = A_i\xx + \ee,
\label{eq:l1min_alignment}
\end{equation}
where $\tau_i\in T$ is in a finite-dimensional group $T$ of transformations
(e.g., affine or homography) acting on the image domain.  The sequence
$\tau_i$ converges to a transformation that aligns the test image $\tilde\bb$
with the training images from the $i$-th class $A_i$. In other words, the algorithm 
extends Lucas-Kanade iterative alignment \cite{LucasB1981} to use the
$\ell_1$-norm as a robust error function, while simultaneously estimating
the coefficients of the linear illumination model $A_i\x$.

Numerically, the original solution \cite{WagnerA2009-CVPR} was implemented based 
on standard interior-point methods, and its speed has been since improved by ALM in a more recent paper
\cite{WagnerA2011-PAMI}. However, due the high per-class cost
of the alignment step, the recognition system still fails to achieve real-time
recognition against datasets of hundreds or thousands of users.  In this paper,
in addition to accelerating the generic
$\ell_1$-min objectives \eqref{eq:l1min} and \eqref{eq:l1min_denoise}, we 
also discuss how to properly accelerate the iterative face alignment step
\eqref{eq:l1min_alignment} on multi-core CPUs and GPUs.

@@

Finally, we present extensive benchmark to compare the performance of the
generic ALM $\ell_1$-min algorithm and the corresponding face recognition
routines on massively parallel CPUs and GPUs.  To this end, a face
recognition system for security and access control applications has been implemented, 
and is diagrammed in Figure \ref{fig:pipeline}. This pipeline follows the work of \cite{WagnerA2011-PAMI}, but
the fast implementations presented in this paper are novel.
Since the parallelization of any algorithm
is highly dependent on the target hardware, we will give a brief overview of
the architecture of our test system in detail in Section
\ref{sec:concurrency}.
Finally, note that 
while this paper focuses on the parallelism available in a single (shared memory) machine,
applications targeting thousands of users or more, will likely require computation be
distributed over a cluster or network of computers, and this is not addressed here.
The source code of our implementations, which target NVIDIA's CUDA GPU architecture
as well as the x86-64 architecture, will be made available for peer evaluation.
\begin{figure}
\centering
{\tiny \input{figures/pipeline_simplified.tex}}
\caption{The face recognition pipeline.}
\label{fig:pipeline}
\end{figure}

@@

Our discussion and experiments will address the most common hardware
configuration for engineering workstations: a motherboard with two quad-core
processors, and a PCI card with a single high-end GPU.  
Recognition involving more than a few hundred users with contemporary hardware
would require a server(or cluster) configuration with an expandable 
number of processors.  We will not address efficient parallelization for 
these systems, which may have additional challenges associated with
their non-shared memory model.\footnote{Non-shared memory configurations
include CPU blade servers, multi-GPU servers, and clusters}.
Unless otherwise specified, all implementations utilize single precision
floating point datatypes.  
% WORKING HERE

@@

\subsection{GPU Hardware Concurrency}
The main defining characteristics of contemporary multi-core GPU architectures
are that they have two (much wider) levels of concurrency, relatively small amounts of
 cache, and relatively low clock speeds.  Whereas most of the transistors on
a typical CPU are dedicated either to cache or to hardware that enables higher
clock speeds (such as branch prediction, out-of-order instruction execution, etc...),
most of the transistors on a GPU are dedicated to arithmetic logic units.
An explanation of the CUDA programming model at a useful level of completeness would take 
more space than is available here, so we will instead frame our discussion in terms of 
hardware capabilities. \footnote{A discussion of the CUDA programming model can be found
in the NVIDIA C for CUDA Programming Guide}

Our target GPU architecture is the NVIDIA Fermi architecture (\eg, the GTX 480
GPU used in this paper).  CUDA based GPUs are comprised of several
\emph{streaming multiprocessors} (SMP), each of which is roughly analogous to a CPU core.
For Fermi architecture GPU's, there are up to 16 SMP's, and each SMP is capable of executing up to 64 single precision 
floating point operations
concurrently. \footnote{In CUDA terms, the warp width is 32, and the floating pipeline can
issue up to two warps simultaneously}

@@

Face alignment \eqref{eq:l1min_alignment} estimates an image transformation
$\tau$ that rectifies the query image $\tilde{\bb}$ with possible pose
variation w.r.t. each training class $A_i$, which leads to the minimal sparsity
in error $\ee$ after the alignment. Note that directly solving
\eqref{eq:l1min_alignment} would require simultaneously estimating $\xx$,
$\ee$, and $\tau$ typically on a nonlinear manifold, which renders the problem
extremely difficult to solve exactly with many local minima.

To mitigate the problem, if the system can obtain a good initial guess of the
transformation $\tau$ (\eg, provided by a good face detector), the optimal
solution for $\tau$ can be sought iteratively by linearization at each $j$th
step as follows:
\begin{equation}
\min_{\xx, \ee, \Delta \tau_j}\|\ee\|_1\quad \subj\quad \tilde\bb\circ\tau_j +  J_j\Delta \tau_j = A_i\xx + \ee,
\end{equation}
where $J_j = \nabla_{\tau_j}(\tilde\bb\circ\tau_j)$ is the Jacobian, and
$\Delta \tau$ is a step update to $\tau$. Denote $\bb_j =
\tilde\bb\circ\tau_j$, $B_j = [A_i, -J_j]$ and $\ww^T = [\xx^T, \Delta
\tau^T]$, then the update $\Delta \tau$ can be computed by solving the
following linear program:
\begin{equation}
\min_{\ww, \ee}\|\ee\|_1\quad \subj\quad \bb_j = B_j\ww + \ee.
\label{eq:alignment-linearization}
\end{equation}
%If the warping $\tau$ is constrained to be a similarity transform,
%it can be paramaterized in such a way that the computation of $\frac{\partial \bb}{\partial \tau_p}$
%reduces to several linear vector operations involving pixel coordinates $(u,v)$, $\tau \in \Re^4$,
%and the image gradients $\f_x, \f_y \in \Re^M$, the latter of which only has to be computed
%once per test image.  
The per-class alignment algorithm via ALM is summarized in Algorithm
\ref{alg:iterative_alignment}. 
%In terms of computational complexity, one significant
%term is the computation of the matrix pseudo-inverse $B_j^\dagger$, whose
%computational cost is bounded by $O(n_i^2m + n_i^3)$. However, for per-class face
%alignment, since both $B_j$ and $J_{j}$ are very tall matrices (\ie, $m\gg n_i$
%for all $i=1, \cdots, K$ classes), we find that the computation cost is still
%dominated by the frequently computed matrix-vector multiplications in the
%inner loop.
\begin{algorithm}[ht!]
\caption{\bf (Face Alignment via ALM)} \label{alg:iterative_alignment}
{\bf Input:} $\bb$, $A_i$, $\x_0 = \mathbf{0}$, $\tau_0$, and $J_0$.
\begin{algorithmic}[1]
\WHILE{not converged ($j = 1,2,\ldots$)}
\STATE Update $\bb_j \leftarrow \frac{\bb\circ \tau_{j-1}}{\|\bb\circ \tau_{j-1}\|}$; $B_j= [A_i, -J_{j-1}]$ and corresponding $(B_j^\dagger)^T$
\STATE Initialize $\ww_0 = \mathbf 0$, $\yy_0 = \mathbf 0$
\WHILE{not converged ($k = 1,2,\ldots$)}
\STATE $\uu_0\leftarrow \ww_{k-1}$; $\zz_0\leftarrow \ee_{k-1}$
\WHILE{not converged ($l = 1,2,\ldots$)}
\STATE $\zz_l \leftarrow \mbox{shrink}\left(\bb_j - B_j\uu_{l-1} + \frac{\yy_{k}}{\mu_{k-1}}, \frac{1}{\mu_{k-1}}\right)$
\STATE $\uu_l \leftarrow B_j^\dagger \left(\bb_j - \zz_{l} + \frac{\yy_{k-1}}{\mu_{k-1}} \right) $
\ENDWHILE
\STATE $\ww_k \leftarrow \uu_l$; $\ee_k \leftarrow \zz_l$
\STATE $\yy_{k} \leftarrow \yy_{k-1} + \mu_{k-1} (\bb_j - B_j\ww_{k} - \e_{k})$
\STATE $\mu_{k} \leftarrow \rho\mu_{k-1}$
\ENDWHILE
\STATE Update $\ee_j$, $\tau_j$, and $J_j$
\ENDWHILE
\end{algorithmic}
{\bf Output:} $\tau_i^*\leftarrow \tau_j, \e_i^*\leftarrow \ee_j$
\end{algorithm}

In summary, the alignment stage essentially contains two levels of available
parallelism. At the higher level, there are per-class alignment problems
that are solved independently, or \emph{problem-level parallelism}.  At a lower
level, the first-order linear algebraic operations exhibit parallelism within
image operations, \ie, at the pixel level.  We call this \emph{pixel-level
parallelism}.  In the following two sections will will discuss how to
exploit these two levels of parallelism on different CPU and GPU architectures.

@@

For the E5530, the L2 cache in each core is only able to store about 16 images.
For image alignment problem, the number of training images per class is
typically larger than 16.  
In this paper we consider two mappings of the parallelism available in the alignment
stage to the concurrency available on the CPU.  We argue that the latter
implementation should achieve higher performance, and later verify that this is the case.

The first implementation disregards 
problem-level parallelism, and instead mapping the pixel-level parallelism
onto both the core-level and the SSE-level concurrency provided by the CPU.  In
other words, alignment is performed of the query
image against the training images of a single user at a time using all cores on
both processors.  The advantage of this implementation is that all of the variables
for the inner loop, most notably $B_j$ and $B_j^\dagger$, fit in L2 cache.

The second implementation maps problem-level parallelism onto core-level
concurrency, and maps pixel-level parallelism onto SSE-level concurrency.  In
other words, eight instances of \eqref{eq:alignment-linearization} are executed
in parallel with each problem running on one of the eight cores.  The advantage
of this implementation is that since the cores are operating on different
alignment problem instances, no data is shared between cores, and therefore no
synchronization is necessary between the cores.  Furthermore, even with eight
problem instances running concurrently, the local variables for the inner
loop still fit in L3 cache.  Thanks to the sequential data access patterns of
the solver, prefetching significantly reduces the bandwidth difference between
L2 and L3 cache.  For these reasons, we expect this implementation to out-perform
the previous implementation.

@@

\subsection{GPU Implementation}
\label{sec:alignment_implementation_gpu}
While on the CPU, algorithm performance is highly dependent on effective
cache usage, on the GPU cache is relatively unimportant for ALM based
$\ell_1$-min.  As can be seen in Figure \ref{fig:caches}, the GPU has 
a very small amount of cache compared to the CPU.  While it might be possible
to fit a single instance of the alignment problem (with a slightly reduced problem size)
into L2 cache, this would not be an efficient use of the GPU's resources.
The strength of the GPU's memory architecture for our purposes is it's ability to sustain a very
high bandwidth to DRAM.  This bandwidth is achieved by having a very large number of
threads issuing memory requests simultaneously.  Solving many alignment problem instances
concurrently increases the number of threads that can be used.

Therefore, our recommended GPU implementation is strikingly similar in spirit
to our recommended CPU implementation: on the CPU we use the cores to solve
multiple instances concurrently, and on the GPU we use multiple SMP's for the 
same purpose.
\footnote{In CUDA terminology, Our proposed alignment stage
implementation consists of a single kernel that performs alignment for all
subject classes.  Each instance of the alignment problem is assigned its own
thread block, and the GPU hardware schedules as many thread blocks to run
concurrently as the hardware will allow.} The number of subject classes that
actually get aligned concurrently is ultimately chosen by the GPU hardware, 
but can be influenced by manual tuning of the code. We have empirically determined
that performance is highest with 5-7 subject classes aligned simultaneously on each SMP.
%This mapping of the problem-level parallelism onto the SMP-level
%concurrency is illustrated in Figure \ref{fig:alignment_mapping_gpu}.
%\begin{figure}
%\centering
%\includegraphics[width=3.4in]{figures/alignment_mapping_gpu}
%\caption{Proposed mapping of alignment parallelism onto GPU concurrency}
%\label{fig:alignment_mapping_gpu}
%\end{figure}

Several other aspects of the implementation merit discussion.  
The full resolution query image, along with its numerical derivatives in both
image directions, are stored in GPU texture memory. This  allows the
computation of $b(\tau)$ and $J(\tau)$, which essentially consist of warped
versions of the test image and its derivatives, to exploit the GPU's special
purpose hardware for bilinear interpolation.

Solving one alignment problem per SMP necessitates the implementation of a
custom CUDA routine for computing $B_j^\dagger = (B_j^TB_j)^{-1} B^T = G^{-1} B_j^T$. $G$
is inverted using Gaussian elimination with partial pivoting.  To achieve
precision comparable to the single precision LAPACK routines in Intel's MKL
with this simplistic algorithm, we use double precision.  Since $G$ is only $n
\times n$ and $B_j^\dagger$ is only computed once per optimization problem, the
cost of the inversion is dwarfed by the cost of other steps.  
Additionally, reductions on long vectors (i.e. sum or $\ell_1$ norm
of an image) are performed using a double precision temporary variable for accumulation.
%In Section \ref{sec:experiment}, we show the performance of our CPU and GPU
%implementations running on synthetic and real image data.

\section{Parallelization of the Face Recognition Stage}
\label{sec:recognition}
After the face alignment stage is complete, all but the 20 users with lowest
$||\e_i||_1$ are thrown out, $\bb$ and $A_i$ are re-sampled to a common alignment
using $\tau_i$, and $A_i$ are concatenated into a new matrix $A$.
A sparse representation of $\bb$ w.r.t $A$ is then recovered
by solving a single $\ell_1$-min problem, as shown
in \eqref{eq:l1min_denoise}, using Algorithm \ref{alg:alm_rec}.  The
coefficients $\x$ are then used to compute error residuals that are used for
classification.  
%This recognition algorithm has
%been extensively studied in
%\cite{WrightJ2009-PAMI,YangA2010-ICIP,WagnerA2011-PAMI}.
In this stage of the algorithm, there is no problem-level
parallelism to exploit, so this section will discuss how to exploit pixel-level parallelism
on both CPU and GPU hardware.
Then in Section \ref{sec:experiment}, we will benchmark the
performance of the two architectures and further demonstrate the speed
gains achieved by our proposed implementations over previously published implementations.

@@

On the GPU, most of the operations map well onto similar calls in NVIDIA's GPU
BLAS library (CUBLAS) which, like MKL, can take advantage of both levels of
concurrency available in the hardware architecture.  Operations that do not map
well onto the CUBLAS API were implemented as GPU kernels (in C for CUDA).
Additionally, some operations that could have been implemented via multiple
BLAS calls, performance improvements were achieved by combining multiple
vector-vector operations into a single kernel, due to reduced bandwidth and
kernel call overhead.

>>>>

\begin{abstract} Recently a family of promising face recognition algorithms
based on sparse representation and $\ell_1$-minimization ($\ell_1$-min) have been
developed.  These algorithms have not yet seen commercial application, 
largely due to higher computational cost compared to other traditional
algorithms. This paper studies techniques for leveraging the 
massive parallelism available in GPU and CPU hardware to accelerate
$\ell_1$-min based on augmented Lagrangian method (ALM) solvers. 
Due to the low arithmetic intensity of ALM solvers, both
architectures are bandwidth limited.  
with large problems running faster on the GPU, and small problems running faster
on the CPU.  
On both architectures, the proposed implementations dramatically out-perform
naive library-based implementations. 
At the problem sizes most important for $\ell_1$-min based
face recognition, both architectures achieve comparable performance.
The source code of the
algorithms will be made available for peer evaluation.  \end{abstract}

@@

Among its many applications, $\ell_1$-min has been recently used to reformulate
image-based face recognition as a sparse representation problem
\cite{WrightJ2009-PAMI}.  If we stack the pixels of the training images of $K$ subject
classes into the columns of matrices $(A_1\in\Re^{d\times n_1}, \cdots, A_K\in\Re^{d\times n_K})$, combine
the matrices into a larger matrix $A = [A_1, \cdots, A_K]\in\Re^{d\times n}$, and arrange the pixels of a new
query image into a vector $\bb\in\Re^d$, \emph{sparsity-based
classification} (SBC) solves the following minimization problem:
\begin{equation}
\min_{\xx, \ee} \| \xx \|_1 + \|\ee\|_1 \quad \subj \quad \bb = A \xx + \ee.
\label{eq:l1min_denoise}
\end{equation}
If the sparsest solutions for $\x$ and $\e$ are recovered, $\ee$ provides a
means to compensate for pixels that are corrupted due to occlusion of some part of the query
image, and the dominant nonzero coefficients of $\xx$ reveal the membership of
$\bb$ based on the training image labels associated with $A$.  Since $A$
contains many images per user taken under different illuminations, 
$A_i\x_i$ acts as a linear illumination model for the test image $\bb$.

In this paper, we study parallelization of $\ell$-min on many-core CPU and GPU
architectures. Although $\ell_1$-min \eqref{eq:l1min} is a convex
program, conventional algorithms such as interior-point methods
\cite{ChenS2001-SIAM,TibshiraniR1996} scale poorly
for applications such as face recognition. Recently, a number of
accelerated algorithms have been proposed that explicitly take advantage of
the special structure of $\ell_1$-min problems
\cite{LorisI2009,YangA2010-ICIP}. We investigate parallelization of a
state-of-the-art $\ell_1$-min solution based on the classical framework of
\emph{augmented Lagrangian methods} (ALM) \cite{BertsekasD2003,YangA2010-ICIP}.

While solving \eqref{eq:l1min_denoise} can achieve exceptional
recognition accuracy on public databases, its success also relies on the assumption
that the query image is well aligned with the training images. In
\cite{WagnerA2009-CVPR}, the framework of SBC was extended to align
a query image to each subject class individually. The face alignment algorithm iteratively solves
the following minimization problem:
\begin{equation}
\hat{\tau}_i = \arg\min_{\xx, \ee, \tau_i} \|\ee\|_1\quad \mbox{subj. to}\quad \tilde\bb\circ\tau_i = A_i\xx + \ee,
\label{eq:l1min_alignment}
\end{equation}
where $\tau_i\in T$ is in a finite-dimensional group $T$ of transformations
(e.g., affine or homography) acting on the image domain.  The sequence
$\tau_i$ converges to a transformation that aligns the test image $\tilde\bb$
with the training images from the $i$-th class $A_i$. In other words, the algorithm 
extends Lucas-Kanade iterative alignment \cite{LucasB1981} to use the
$\ell_1$-norm as a robust error function, while simultaneously estimating
the coefficients of the linear illumination model $A_i\x$.

Numerically, the original solution \cite{WagnerA2009-CVPR} was implemented based 
on standard interior-point methods, and its speed has been since improved by ALM in a more recent paper
\cite{WagnerA2011-PAMI}. However, due the high per-class cost
of the alignment step, the recognition system still fails to achieve real-time
recognition against datasets of hundreds or thousands of subjects.  In this paper,
in addition to accelerating the generic
$\ell_1$-min objectives \eqref{eq:l1min} and \eqref{eq:l1min_denoise}, we 
also discuss how to efficiently accelerate the iterative face alignment step
\eqref{eq:l1min_alignment} on multi-core CPUs and GPUs.

@@

Finally, we present extensive benchmark to compare the performance of the
generic ALM $\ell_1$-min algorithm and the corresponding face recognition
routines on massively parallel CPUs and GPUs.  To this end, a face
recognition system for security and access control applications has been implemented, 
and is diagrammed in Figure \ref{fig:pipeline}. This pipeline follows the work of \cite{WagnerA2011-PAMI}, but
the fast implementations presented in this paper are novel.
Since the parallelization of any algorithm
is highly dependent on the target hardware, we will give a brief overview of
the architecture of our test system in detail in Section
\ref{sec:concurrency}.
Finally, note that 
while this paper focuses on the parallelism available in a single (shared memory) machine,
applications targeting thousands of subjects or more, will likely require computation be
distributed over a cluster or network of computers, and this is not addressed here.
The source code of our implementations, which target NVIDIA Fermi GPUs
as well as Intel x86-64 CPUs, will be made available for peer evaluation.
\begin{figure}
\centering
{\tiny \input{figures/pipeline_simplified.tex}}
\caption{The face recognition pipeline.}
\label{fig:pipeline}
\end{figure}

@@

Our discussion and experiments will address the most common hardware
configuration for engineering workstations: a motherboard with two quad-core
processors, and a PCI card with a single high-end GPU.  
Recognition involving more than a few hundred subjects with contemporary hardware
would require a server(or cluster) configuration with an expandable 
number of processors.  We will not address efficient parallelization for 
these systems, which may have additional challenges associated with
their non-shared memory model.\footnote{Non-shared memory configurations
include CPU blade servers, multi-GPU servers, and clusters}.
Unless otherwise specified, all implementations utilize single precision
floating point datatypes.  
% WORKING HERE

@@

\subsection{GPU Hardware Concurrency}
The main defining characteristics of contemporary multi-core GPU architectures
are that they have two (much wider) levels of concurrency, relatively small amounts of
 cache, and relatively low clock speeds.  Whereas most of the transistors on
a typical CPU are dedicated either to cache or to hardware that enables higher
clock speeds (such as branch prediction, out-of-order instruction execution, etc...),
most of the transistors on a GPU are dedicated to arithmetic logic units.
For our GPU implementations we target NVIDIA Fermi GPU architecture (\eg, the GTX 480
GPU used in this paper).  An explanation of the GPU programming model (CUDA) at a useful level of completeness would take more space than is available here, so we will instead frame our discussion in terms of 
hardware capabilities. \footnote{A discussion of the CUDA programming model can be found
in the NVIDIA C for CUDA Programming Guide}
CUDA programmable GPUs are comprised of several
\emph{streaming multiprocessors} (SMP), each of which is roughly analogous to a CPU core.
For Fermi architecture GPU's, there are up to 16 SMP's, and each SMP is capable of executing up to 64 single precision 
floating point operations
concurrently. \footnote{In CUDA terms, the warp width is 32, and the floating pipeline can
issue up to two warps simultaneously}

@@

Face alignment \eqref{eq:l1min_alignment} estimates an image transformation
$\tau$ that rectifies the query image $\tilde{\bb}$ with possible pose
variation w.r.t. each training class $A_i$, which leads to the minimal sparsity
in error $\ee$ after the alignment. Note that directly solving
\eqref{eq:l1min_alignment} is impossible since it is a non-convex problem
and may exhibit local minima.
However, given a good initial guess of the
transformation $\tau$ (\eg, provided by an accurate face detector), the optimal
solution for $\tau$ can be sought iteratively by linearization at each $j$th
step:
\begin{equation}
\min_{\xx, \ee, \Delta \tau_j}\|\ee\|_1\quad \subj\quad \tilde\bb\circ\tau_j +  J_j\Delta \tau_j = A_i\xx + \ee,
\end{equation}
where $J_j = \nabla_{\tau_j}(\tilde\bb\circ\tau_j)$ is the Jacobian, and
$\Delta \tau$ is a step update to $\tau$. Denote $\bb_j =
\tilde\bb\circ\tau_j$, $B_j = [A_i, -J_j]$ and $\ww^T = [\xx^T, \Delta
\tau^T]$, then the update $\Delta \tau$ can be computed by solving the
following linear program:
\begin{equation}
\min_{\ww, \ee}\|\ee\|_1\quad \subj\quad \bb_j = B_j\ww + \ee.
\label{eq:alignment-linearization}
\end{equation}
%If the warping $\tau$ is constrained to be a similarity transform,
%it can be paramaterized in such a way that the computation of $\frac{\partial \bb}{\partial \tau_p}$
%reduces to several linear vector operations involving pixel coordinates $(u,v)$, $\tau \in \Re^4$,
%and the image gradients $\f_x, \f_y \in \Re^M$, the latter of which only has to be computed
%once per test image.  
The per-class alignment algorithm via ALM is summarized in Algorithm
\ref{alg:iterative_alignment}. 
%In terms of computational complexity, one significant
%term is the computation of the matrix pseudo-inverse $B_j^\dagger$, whose
%computational cost is bounded by $O(n_i^2m + n_i^3)$. However, for per-class face
%alignment, since both $B_j$ and $J_{j}$ are very tall matrices (\ie, $m\gg n_i$
%for all $i=1, \cdots, K$ classes), we find that the computation cost is still
%dominated by the frequently computed matrix-vector multiplications in the
%inner loop.
\begin{algorithm}[ht!]
\caption{\bf (Face Alignment via ALM)} \label{alg:iterative_alignment}
{\bf Input:} $\bb$, $A_i$, $\x_0 = \mathbf{0}$, $\tau_0$, and $J_0$.
\begin{algorithmic}[1]
\WHILE{not converged ($j = 1,2,\ldots$)}
\STATE Update $\bb_j \leftarrow \frac{\bb\circ \tau_{j-1}}{\|\bb\circ \tau_{j-1}\|}$; $B_j= [A_i, -J_{j-1}]$ and corresponding $(B_j^\dagger)^T$
\STATE Initialize $\ww_0 = \mathbf 0$, $\yy_0 = \mathbf 0$
\WHILE{not converged ($k = 1,2,\ldots$)}
\STATE $\uu_0\leftarrow \ww_{k-1}$; $\zz_0\leftarrow \ee_{k-1}$
\WHILE{not converged ($l = 1,2,\ldots$)}
\STATE $\zz_l \leftarrow \mbox{shrink}\left(\bb_j - B_j\uu_{l-1} + \frac{\yy_{k}}{\mu_{k-1}}, \frac{1}{\mu_{k-1}}\right)$
\STATE $\uu_l \leftarrow B_j^\dagger \left(\bb_j - \zz_{l} + \frac{\yy_{k-1}}{\mu_{k-1}} \right) $
\ENDWHILE
\STATE $\ww_k \leftarrow \uu_l$; $\ee_k \leftarrow \zz_l$
\STATE $\yy_{k} \leftarrow \yy_{k-1} + \mu_{k-1} (\bb_j - B_j\ww_{k} - \e_{k})$
\STATE $\mu_{k} \leftarrow \rho\mu_{k-1}$
\ENDWHILE
\STATE Update $\ee_j$, $\tau_j$, and $J_j$
\ENDWHILE
\end{algorithmic}
{\bf Output:} $\tau_i^*\leftarrow \tau_j, \e_i^*\leftarrow \ee_j$
\end{algorithm}

In summary, the alignment stage essentially contains two levels of available
parallelism. At the higher level, there are per-class alignment problems
that are solved independently, or \emph{problem-level parallelism}.  At a lower
level, the first-order linear algebraic operations exhibit parallelism within
image operations, \ie, at the pixel level.  We call this \emph{pixel-level
parallelism}.  The following two sections discuss methods for exploiting
these two levels of parallelism on CPU and GPU architectures.

@@

For the E5530, the L2 cache in each core is only able to store about 16 images.
For image alignment problem, the number of training images per class is
typically larger than 16.  
In this paper we compare two mappings of the parallelism available in the alignment
stage to the concurrency available on the CPU: a naive implementation that is purely 
library-based, and the optimized version we advocate.

The naive implementation disregards 
problem-level parallelism, and instead maps pixel-level parallelism
onto both the core-level and the SSE-level concurrency provided by the CPU.  In
other words, alignment is performed of the query
image against the training images of a single user at a time using all cores on
both processors.  The potential advantage of this implementation is that all of the variables
for the inner loop, most notably $B_j$ and $B_j^\dagger$, fit in L2 cache.

The optimized implementation maps problem-level parallelism onto core-level
concurrency, and maps pixel-level parallelism onto SSE-level concurrency.  In
other words, eight instances of \eqref{eq:alignment-linearization} are executed
in parallel with each problem running on one of the eight cores.  The advantage
of this implementation is that since the cores are operating on different
alignment problem instances, no data is shared between cores, and therefore no
synchronization is necessary between the cores.  Furthermore, even with eight
problem instances running concurrently, the local variables for the inner
loop still fit in L3 cache.  Because of the sequential data access patterns of
the solver, the CPU hardware is able to use the full L3 cache bandwidth, which is 
high enough to make the program CPU limited, rather than memory limited.  
For these reasons, we expect this implementation to out-perform
the previous implementation.

@@

\subsection{GPU Implementation}
\label{sec:alignment_implementation_gpu}
While on the CPU, algorithm performance is highly dependent on effective
cache usage, on the GPU cache is relatively unimportant for ALM based
$\ell_1$-min.  As can be seen in Figure \ref{fig:caches}, the GPU has 
a very small amount of cache compared to the CPU.  While it might be possible
to fit a single instance of the alignment problem (with a slightly reduced problem size)
into L2 cache, this would not be an efficient use of the GPU's resources.
The strength of the GPU's memory architecture for our purposes is it's ability to sustain a very
high bandwidth to DRAM.  This bandwidth is achieved by having a very large number of
threads issuing interleaved memory requests.  Solving many alignment problem instances
concurrently increases the number of threads that can be used.

Therefore, our recommended GPU implementation is strikingly similar in spirit
to our recommended CPU implementation: on the CPU we use the cores to solve
multiple instances concurrently, and on the GPU we use multiple SMP's for the 
same purpose.
\footnote{In CUDA terms, our proposed alignment stage
implementation consists of a single kernel that performs alignment for all
subject classes.  Each instance of the alignment problem is assigned its own
thread block, and the GPU hardware schedules as many thread blocks to run
concurrently as the hardware will allow.} The number of subject classes that
are actually scheduled to run concurrently is chosen by the GPU hardware, 
but can be indirectly influenced by manual tuning of the code.\footnote{In CUDA terms, it is the kernel launch
configuration that is being tuned.} We have empirically determined
that performance is highest with 5-7 subject classes aligned simultaneously on each SMP.
%This mapping of the problem-level parallelism onto the SMP-level
%concurrency is illustrated in Figure \ref{fig:alignment_mapping_gpu}.
%\begin{figure}
%\centering
%\includegraphics[width=3.4in]{figures/alignment_mapping_gpu}
%\caption{Proposed mapping of alignment parallelism onto GPU concurrency}
%\label{fig:alignment_mapping_gpu}
%\end{figure}

Several other aspects of the implementation merit discussion.  
First of all, we take advantage of the GPU's special hardware dedicated to bilinear interpolation
for the computation of $b(\tau)$ and $J(\tau)$, which essentially consist of resampled
versions of the test image and its derivatives.
Second of all, since there are no standard CUDA libraries that work at the SMP level, we use a
custom routine for computing $B_j^\dagger = (B_j^TB_j)^{-1} B^T = G^{-1} B_j^T$. $G$
is inverted using Gaussian elimination with partial pivoting.  To achieve
precision comparable to the single precision LAPACK routines in Intel's MKL
with this simplistic algorithm, we use double precision.  Since $G$ is only $n
\times n$ and $B_j^\dagger$ is only computed once per optimization problem, the
cost of the inversion is dwarfed by the cost of other steps.  
Similarly, reductions (i.e. sum or $\ell_1$ norm) on image-sized vectors are also
performed in double precision.
%The lack of a atomic floating point addition on the GPU
%significantly reduces the perfomance of this step of the algorithm.
%In Section \ref{sec:experiment}, we show the performance of our CPU and GPU
%implementations running on synthetic and real image data.

\section{Parallelization of the Face Recognition Stage}
\label{sec:recognition}
After the face alignment stage is complete, the 20 subjects with lowest
$||\e_i||_1$ are selected for recognition, $\bb$ and $A_i$ are re-sampled to a common alignment
using $\tau_i$, and $A_i$ are concatenated into a new matrix $A$.
A sparse representation of $\bb$ w.r.t $A$ is then recovered
by solving a single $\ell_1$-min problem, as shown
in \eqref{eq:l1min_denoise}, using Algorithm \ref{alg:alm_rec}.  The
coefficients $\x$ are then used to compute error residuals that are used for
classification.  
%This recognition algorithm has
%been extensively studied in
%\cite{WrightJ2009-PAMI,YangA2010-ICIP,WagnerA2011-PAMI}.
In this stage of the algorithm, there is no problem-level
parallelism to exploit, so this section will discuss how to exploit pixel-level parallelism
on both CPU and GPU hardware.
Then in Section \ref{sec:experiment}, we will benchmark the
performance of the two architectures and further demonstrate the speed
gains achieved by our proposed implementations over previously published implementations.

@@

On the GPU, most of the operations map well onto similar calls in NVIDIA's GPU
BLAS library (CUBLAS) which, like MKL, can take advantage of both levels of
concurrency available in the hardware architecture.  Operations that do not map
well onto the CUBLAS API were implemented directly in CUDA code.
Additionally, some operations that could have been implemented via multiple
BLAS calls, performance improvements were achieved by combining multiple
vector-vector operations into a single kernel, due to reduced bandwidth and
kernel call overhead.

In order to avoid expensive data transfer across the bus connecting the GPU
card and the CPU motherboard (the PCI express bus), all of the data is
transferred to GPU DRAM once, and all non-trivial tasks are performed on the
GPU on data stored in GPU DRAM.  
%with the sole exception of the pseudo-inverse of $B_j$,
%which is computed on the CPU via Intel's LAPACK implementation, and then
%uploaded to GPU DRAM.
